A contrast of meta and metafor packages for meta-analyses in R

### 2020-03-27 ms/2020/contrastRmd.Rmd
tidy up

---
title: A contrast of meta and metafor packages for meta-analyses in R
preprint: true
author: 
  - name: Christopher J. Lortie
    affiliation: 1
    corresponding: true
    email: chris@christopherlortie.info

affiliation:
  - code: 1
    address: NCEAS, UCSB, Santa Barbara, CA. 93101

abstract: >
  There is extensive support and choice in R to support meta-analyses.  Two common packages in the natural sciences include meta and metafor.  Here, a brief contrast of the strengths of each is described for the synthesis scientist.  Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.  Metafor is a comprehensive package for analyses if the fit models are more complex.  Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.  Preference and critical outcomes can facilitate choice between these two specific options.  Nonetheless, metafor has a steeper learning curve but greater rewards.
header-includes: >
  \usepackage{lipsum}
bibliography: references.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

# Introduction {-}

Meta-analyses are common and powerful synthesis tools in science.  Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications [@RN3629, @RN4850].  A meta-analysis in the natural sciences, i.e. outside the health sciences, is comprised of a formalized systematic review and analysis of effect sizes and is termed a meta-analysis when statistics examining intervention efficacy are included [@RN3216].  In other fields, the terms systematic review and meta-analysis are used more interchangeable, and meta-statistics are often done on compiled randomized controlled trials or other relatively large datasets in addition to sets of data derived from peer-reviewed publications.  The derived data in the natural sciences (as the intended primary audience here) routinely extracts only the mean values or single point measures from publications [@RN4861].  The synthesis statistics were commonly done using MetaWin [@RN6194] or other GUI-based applications for a number of years.  More recently however, statistics in the fields of ecology and evolution for instance have increasingly moved to the programming language R [@RN6098], and synthesis statistics are no exception.  At least two R packages have risen to prominence for general meta-analytical statistics in the natural sciences - namely meta [@RN6176] and metafor [@RN6175].  Given that meta-analyses are also increasingly published in these same fields [@RN2189, @RN3629], a brief comment on the ecosystem of analytical choices that R provides is beneficial and timely.  We need synthesis to inform evidence-based decisioning, and meta-analyses can be the primary tool if aggregated primary datasets are unavailable.  Furthermore even with primary data in hand, data reduction to effect sizes within primary and synthesis studies is a mechanism to illustrate differences and strength of effects.  These approaches provide the capacity for higher-order analyses and reuse [@RN4835] suggesting that familiarity with effect sizes is both germane and practical.

# The R ecosystem for meta-analyses {-}
Like many fundamental challenges in science, the R developer community provides potential solution sets distributed across multiple packages for synthesis.  Broadly speaking, alternative packages in R sometimes examine an issue from different perspectives and provide unique functions.  In other instances, packages can be very similar or analogs in terms of functionality and use conceptually aligned functions that differ only in nomenclature or arguments.  Scientific synthesists that choose to do a meta-analyses in R have options.  A total of 63 packages associated with various aspects of conducting a meta-analysis have been identified in a comprehensive review and typology of options [@RN6178].  Both meta [@RN6176] and metafor [@RN6175] are amongst 11 generic packages identified [@RN6178].  These two packages are analogs but with different inherent workflows.  There is also rmeta for simple fixed and random effects meta-analyses [@RN6195], mada for diagnostics [@RN6196], [@RN6196], netmeta for frequentist network meta statistics [@RN6197], and mvmeta for multivariate derived data aggregations [@RN6198] to name a few options.  The latter three packages listed have distinct and specific niches for analysis whilst meta and metafor overlap considerably.  Consequently, a brief contrast in facilitating choice between these two packages for the general analyst is provided here.

# Contrast of meta versus metafor {-}
Meta is a well-maintained, recently updated CRAN R package (Version 4.9-4 updated on January 3, 2019) with 31 unique functions, 7 sample datasets, and a reference manual.  There is also an exceptional textbook devoted to meta-analysis in R that focuses primarily on this package [@RN6199].  It is highly capable of resolving most general meta-analytical challenges that an analyst will face including the capacity to include Empirical Bayes estimators as arguments in some functions, predictive meta-statistics, interaction terms, meta-regression, and modifiers.  Note that for some of these methods, the rma.uni function is sourced internally from the package metafor.  This is intriguing but mostly opaque and inconsequential to the user if she prefers the structure of the arguments within functions, the semantics, or the workflow of the meta package.  The primary strengths include its direct and straightforward implementation with minimal (source) lines of code to do an analysis.  Provided one has secured the derived data from the studies and organized into a dataframe with vectors as each key argument within the main meta-model fitting functions, statistics are simple.  The type of response variable such as mean, continuous, or rate is matched to a specific function call such as metamean, metacont, or metarate.  This is semantically intuitive and encourages good thinking before statistics because it engenders consideration of the data.  The effect size calculation is included in this main function and defaults return the most prevalent effect size measure typically associated with those data, but it can also be specified as an argument.  The primary workflow is thus a single step if the user elects to rely on the internal calculations provided in this package.  Exploration of the model is well articulated with funnel, radial, and forest plots.  Z-scores, significance tests, and heterogeneity statistics are printed in the model summary. Publication bias is also provided as a more in-depth function entitled metabias within this package.  There are two standout functions in this package.  The first is a function entitled metagen, and it is a backup, multipurpose tool so to speak that fits a generic inverse variance meta-analysis.  This a handy tool for user-calculated effect size measures or for exploration of statistical trends with reduced data assumptions.  In some fields, there are specific effect size estimates that this function provides a robust, easy-to-fit capacity for statistics.  The second standout function is bubble.metareg for a quick, visual exploration of the outcome of a meta-regression.  It is useful in contemporary data science to use visualization as a means to understand data [@RN4510], but statistical packages do not always provide the means to easily iterate between statistics or model fitting and visualization.  In summary, excepting unique data or statistical issues, this package is directly implemented and effective.

@@

# Conclusions {-}

Statistics are sometimes about preferences and thinking styles [@RN6087], and scientific synthesis is both an art and a science [@RN3629].  Trade-offs are also common in adopting one ecosystem, analysis tool, or specific package for data wrangling and analyses.  If more rapid, less specified, general meta-analyses are the goal – the package meta is a direct means to an end.  Moderators are added post hoc in additional, update model steps, but the first model fit is a single, intuitive process.  Meta-regression is viable and interaction terms can be included.  The generic meta-analysis function is a superb tool.  Metafor requires the effect size compilation a priori and is thus a bit more coding to prepare for the meta-model.  However, deeper and more complex model fits are inherent in the semantics of these functions.  If the synthesist does have not effect size measure in hand or wishes to calculate effect sizes measures but not for meta-models, the escalc function is invaluable in this package.  In summary, both packages provide the capacity for basic and advanced meta-analyses but more advanced modelling is likely worth the commitment to metafor.

# Acknowledgments {-}  
CJL is a Senior Research Associate at NCEAS and in-kind computational support is provided.

# References

>>>>

---
title: "A contrast of meta and metafor packages for meta-analyses in R"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Christopher J. Lortie

Department of Biology, York University  
Toronto, ON, Canada.  M3J1P3.  
[lortie@yorku.ca](mailto:lortie@yorku.ca)  
PH: 416.736.2100 x20588

### Abstract
There is extensive choice in R to support meta-analyses.  Two common packages in the natural sciences include meta and metafor.  A brief contrast of the strengths of each is described here for the synthesis scientist.  Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.  Metafor is a comprehensive package for analyses for relatively more complex fit models.  Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.  Preference and critical outcomes can facilitate choice between these two specific options.  Nonetheless, metafor has a steeper learning curve but greater rewards.

\newpage

### Introduction
Meta-analyses are common and powerful synthesis tools in science.  Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications [@RN3629, @RN4850].  A meta-analysis in the natural sciences, i.e. outside the health sciences, is comprised of a formalized systematic review and analysis of effect sizes and is termed a meta-analysis when statistics examining intervention efficacy are included [@RN3216].  In other fields, the terms systematic review and meta-analysis are used more interchangeable, and meta-statistics are often done on compiled randomized controlled trials or other relatively large datasets in addition to sets of data derived from peer-reviewed publications.  The derived data in the natural sciences (as the intended primary audience here) routinely extracts only the mean values or single point measures from publications [@RN4861].  The synthesis statistics were commonly done using MetaWin [@RN6194] or other GUI-based applications for a number of years.  More recently however, statistics in the fields of ecology and evolution for instance have increasingly moved to the programming language R [@RN6098], and synthesis statistics are no exception.  At least two R packages have risen to prominence for general meta-analytical statistics in the natural sciences - namely meta [@RN6176] and metafor [@RN6175].  Given that meta-analyses are also increasingly published in these same fields [@RN2189, @RN3629], a brief comment on the ecosystem of analytical choices that R provides is beneficial and timely.  We need synthesis to inform evidence-based decisioning, and meta-analyses can be the primary tool if aggregated primary datasets are unavailable.  Furthermore even with primary data in hand, data reduction to effect sizes within primary and synthesis studies is a mechanism to illustrate differences and strength of effects.  These approaches provide the capacity for higher-order analyses and reuse [@RN4835] suggesting that familiarity with effect sizes is both germane and practical.

### The R ecosystem for meta-analyses  
Like many fundamental challenges in science, the R developer community provides potential solution sets distributed across multiple packages for synthesis.  Broadly speaking, alternative packages in R sometimes examine an issue from different perspectives and provide unique functions.  In other instances, packages can be very similar or analogs in terms of functionality and use conceptually aligned functions that differ only in nomenclature or arguments.  Scientific synthesists that choose to do a meta-analyses in R have options.  A total of 63 packages associated with various aspects of conducting a meta-analysis have been identified in a comprehensive review and typology of options [@RN6178].  Both meta [@RN6176] and metafor [@RN6175] are amongst 11 generic packages identified [@RN6178].  These two packages are analogs but with different inherent workflows.  There is also rmeta for simple fixed and random effects meta-analyses [@RN6195], mada for diagnostics [@RN6196], [@RN6196], netmeta for frequentist network meta statistics [@RN6197], and mvmeta for multivariate derived data aggregations [@RN6198] to name a few options.  The latter three packages listed have distinct and specific niches for analysis whilst meta and metafor overlap considerably.  Consequently, a brief contrast in facilitating choice between these two packages for the general analyst is provided here.

### Contrast of meta versus metafor  
Meta is a well-maintained, recently updated CRAN R package (Version 4.9-4 updated on January 3, 2019) with 31 unique functions, 7 sample datasets, and a reference manual.  There is also an exceptional textbook devoted to meta-analysis in R that focuses primarily on this package [@RN6199].  It is highly capable of resolving most general meta-analytical challenges that an analyst will face including the capacity to include Empirical Bayes estimators as arguments in some functions, predictive meta-statistics, interaction terms, meta-regression, and modifiers.  Note that for some of these methods, the rma.uni function is sourced internally from the package metafor.  This is intriguing but mostly opaque and inconsequential to the user if she prefers the structure of the arguments within functions, the semantics, or the workflow of the meta package.  The primary strengths include its direct and straightforward implementation with minimal (source) lines of code to do an analysis.  Provided one has secured the derived data from the studies and organized into a dataframe with vectors as each key argument within the main meta-model fitting functions, statistics are simple.  The type of response variable such as mean, continuous, or rate is matched to a specific function call such as metamean, metacont, or metarate.  This is semantically intuitive and encourages good thinking before statistics because it engenders consideration of the data.  The effect size calculation is included in this main function and defaults return the most prevalent effect size measure typically associated with those data, but it can also be specified as an argument.  The primary workflow is thus a single step if the user elects to rely on the internal calculations provided in this package.  Exploration of the model is well articulated with funnel, radial, and forest plots.  Z-scores, significance tests, and heterogeneity statistics are printed in the model summary. Publication bias is also provided as a more in-depth function entitled metabias within this package.  There are two standout functions in this package.  The first is a function entitled metagen, and it is a backup, multipurpose tool so to speak that fits a generic inverse variance meta-analysis.  This a handy tool for user-calculated effect size measures or for exploration of statistical trends with reduced data assumptions.  In some fields, there are specific effect size estimates that this function provides a robust, easy-to-fit capacity for statistics.  The second standout function is bubble.metareg for a quick, visual exploration of the outcome of a meta-regression.  It is useful in contemporary data science to use visualization as a means to understand data [@RN4510], but statistical packages do not always provide the means to easily iterate between statistics or model fitting and visualization.  In summary, excepting unique data or statistical issues, this package is directly implemented and effective.

@@

### Conclusions  
Statistics are sometimes about preferences and thinking styles [@RN6087], and scientific synthesis is both an art and a science [@RN3629].  Trade-offs are also common in adopting one ecosystem, analysis tool, or specific package for data wrangling and analyses.  If more rapid, less specified, general meta-analyses are the goal – the package meta is a direct means to an end.  Moderators are added post hoc in additional, update model steps, but the first model fit is a single, intuitive process.  Meta-regression is viable and interaction terms can be included.  The generic meta-analysis function is a superb tool.  Metafor requires the effect size compilation a priori and is thus a bit more coding to prepare for the meta-model.  However, deeper and more complex model fits are inherent in the semantics of these functions.  If the synthesist does have not effect size measure in hand or wishes to calculate effect sizes measures but not for meta-models, the escalc function is invaluable in this package.  In summary, both packages provide the capacity for basic and advanced meta-analyses but more advanced modelling is likely worth the commitment to metafor.

\newpage

### Literature cited

### 2020-03-27 ms/lortie_meta/lortie_meta.Rmd
tidy up

abstract: >
  There is extensive support and choice in R to support meta-analyses.  Two common packages in the natural sciences include meta and metafor.  Here, a brief contrast of the strengths of each is described for the synthesis scientist.  Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.  Metafor is a comprehensive package for analyses if the fit models are more complex.  Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.  Preference and critical outcomes can facilitate choice between these two specific options.  Nonetheless, metafor has a steeper learning curve but greater rewards.
header-includes: >
  \usepackage{lipsum}
bibliography: references.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

>>>>

abstract: >
  There is extensive choice in R to support meta-analyses.  Two common packages in the natural sciences include meta and metafor.  A brief contrast of the strengths of each is described here for the synthesis scientist.  Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.  Metafor is a comprehensive package for analyses for relatively more complex fit models.  Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.  Preference and critical outcomes can facilitate choice between these two specific options.  Nonetheless, metafor has a steeper learning curve but greater rewards.
header-includes: >
  \usepackage{lipsum}
bibliography: references.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---
